Isomorphic Interpreters from Logically Reversible Abstract Machines

### 2012-02-24 iso-int/rc-iso-int.tex
started rc cleanup

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Simple Bounded Iterator}
Consider definition for the following small-step abstract machine:

%subcode{bnf} include main
% Expressions, e = n
% Numbers, n, m = 0 | n + 1
%
% Machine states = <e, e>
% Start state = <n, 0>
% Stop State = <0, n>

@@

This machine start with a number {{n}} in the first position and at
each reduction step decrements the first number and increments the
second. It stops when the first number reaches 0, thereby taking
exactly {{n}} steps. While the machine does nothing interesting,
it useful to illustrate the general idea.

\begin{enumerate}
\item 
We can turn this abstract machine into a an isomorphic interpreter
implemented in {{langRevT}}, in a systematic way. To start with we
realize that the interpreter should implement the following map.

@@

Now numbers can be implemented by the recursive type {{nat = @@x.1+x}}
in {{langRevT}}. This gives us the isomorphisms

\begin{center}
  {{unfold : nat <-> 1 + nat : fold}}
\end{center}

We start by examining the first value.

\begin{center}
  \includegraphics{diagrams/nat-nat2.pdf}
\end{center}

\item
The triangle denotes the {{unfold}} isomorphism, which works as
follows: If the number {{n}} is zero, the output is the top branch
which has type {{1}}. If the number was non-zero, the output is on the
bottom branch and has value {{n-1}}.

Since the abstract machine increments second component, we know that
we need a {{fold}} on the second component to accomplish this. So we
can fill in the output part of the interpreter dually.

@@

\item
Let us examine all the possible input states, by distributing on the
input. Dually let us do the same with the output, but this time also
be a little bit explicit about the swap operations required so the
right types are in place.

@@

\item
The construction above already gives us some insight into how we can
implement our one reduction step. If we chose {{m = m'-1}} and we
chose {{n-1 =n'}} then we have essentially encoded the required
rewrite. Lets state this in words - the new value of {{n}}, namely
{{n'}}, is equal to {{n-1}}. And further - the new value of {{m}},
namely {{m'}}, is {{m+1}} (because {{m'-1=m}}).

@@

\item
We are a few steps away from an interpreter at this point and the
final steps rely on the answers to a few questions.

@@

\item What do we do about the two inner branches?

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Simple Bounded Iterator}
We illustrate the main ideas underlying the derivation of isomorphic
interpreters using a simple abstract machine.  Consider the definition
of the following small-step abstract machine:

\begin{multicols}{2}
%subcode{bnf} include main
% Numbers, n, m = 0 | n + 1
% Machine states = <n, n>

%subcode{bnf} include main
% Start state = <n, 0>
% Stop State = <0, n>
\end{multicols}
% Small-step operational semantics:

@@

This machine starts with a number {{n}} in the first position and
{{0}} in the second. Each reduction step decrements the first number
and increments the second. The machine stops when the first number
reaches 0, thereby taking exactly {{n}} steps.  
%% We can turn this
%% abstract machine into a an isomorphic interpreter implemented in
%% {{langRevT}}, in a systematic way. 
Broadly, the interpreter should implement the map on the left:

@@

%% \begin{center}
%%   \includegraphics{diagrams/nat-nat2.pdf}
%% \end{center}  
%% \end{multicols}

In {{langRevT}} we have {{nat}} given by the recursive definition
{{@@x.1+x}} along with isomorphisms {{unfold:nat <->1+nat:fold}}. The
only reduction step of the abstract machine examines the first value
of the input and reconstructs the second.

@@

The triangle denotes the {{unfold}} isomorphism, which works as
follows: If the number {{n}} is zero, the output is the top branch
which has type {{1}}. If the number was non-zero, the output is on the
bottom branch and has value {{n-1}}.

%% Now numbers can be implemented by the recursive type {{nat = @@x.1+x}}
%% in {{langRevT}}. This gives us the isomorphisms

%% \begin{center}

%% \end{center}

%% Since the abstract machine increments second component, we know that
%% we need a {{fold}} on the second component to accomplish this. So we
%% can fill in the output part of the interpreter dually.

Let us examine all the possible input states, by distributing on the
input. Dually let us do the same with the output, but this time also
be a little bit explicit about the swap operations required so the
right types are in place.

@@

The construction above already gives us some insight into how we can
implement our one reduction step. If we chose {{m = m'-1}} and we
chose {{n-1 =n'}} then we have essentially encoded the required
rewrite. Lets state this in words - the new value of {{n}}, namely
{{n'}}, is equal to {{n-1}}. And further - the new value of {{m}},
namely {{m'}}, is {{m+1}} (because {{m'-1=m}}).

@@

We are a few steps away from an interpreter at this point and the
final steps rely on the answers to a few questions.

@@

\item

 What do we do about the two inner branches?
